%!TEX root = ../../dissertacao.tex
If computers were a broadly accepted tool for improving work and life, then computer networks were a turning point for human society. Finances, administrative and bureaucratic processes, commercial transactions, social interactions, everything seems to be slowly migrating to a computer system environment, not to mention that almost everything nowadays can be network connected: mobile phones, televisions and even home appliances, like a refrigerator. At a certain point, sensible data and critical information was being handled and, quickly, a new necessity emerged: security.

\textit{Internet banking} is one of these areas, where sensible data, like monetary transactions and users credentials, is a desired target by malicious peers, who try to steal, manipulate or sniff on these data, potentially leading to great monetary or corporations' image damage. As a counter measure, researchers and corporate organizations try to develop ways for providing security for users' property and information through security protocols.

The \textit{Dynamic Authentication Protocol} (DAP)~\cite{Peotta2012} is a security protocol, developed in 2012, aiming to provide a trustful authorization method for each banking transaction on an out-of-band secure access model. The author claims that the protocol provides minimization of success to a vast field of known attacks and approaches. However, the cited document does not provide a formal and incontestable proof of the protocol correctness, only providing to the reader a study case of a real world situation.

\textit{Formal verification} is the field of Computer Science where mathematical tools are used to proving or disproving the correctness of algorithms, software or hardware. In essence, this requires the translation of those systems and their properties into a formal language for further verification by formal methods. In contrast to informal methods, which tend to be less arduous, formal verification underlies in mathematical principles and provides reproducible proofs or counter-examples of a software correctness.

Like a regular computer program, security protocols are software, often involving concurrency, where general scenarios gather multiple peers communicating with each other. Numerous techniques were developed for formally verify security protocols. The \textit{Inductive Method}, first proposed by L.~Paulson~\cite{Paulson98} and further developed by G.~Bella~\cite{Bella2007} is an interesting system, which abstracts many details of a security protocol context in a manner where it is possible to reason about past executions in a formal and logical way.

In this work, the Inductive Method was explored and putted to test on DAP, a real world and challenging protocol, where the chosen area of interest shows significant relevance. Based in techniques from previous works~\cite{BellaPaulson2006, Paulson99, Paulson98}, the protocol was formally specified and verified.

The main results were a reliable formalization of smartphones and its related events of inputing and outputting data and, using such models, the formalization of the DAP. From the formalization, several lemmas regarding the model properties were derived, guaranteeing its reliability to the protocol and entities involved. Additionally, some of the protocol claimed properties were explored and further certified through formal reasoning.




\section{Motivation}
Authentication is the biggest concern in security for internet banking~\cite{Hutchinson2003}. Allied to confidentiality, these properties provides a reliable environment for users to perform monetary transactions without worrying with passive network eavesdroppers or active attackers trying to modify these transactions.

However, despite the efforts to provide desired security qualities for internet banking security systems, criminals are still able to perform successful attacks on these infrastructures~\cite{Adham2013}.

Promising works for ensuring security have appeared over years using OTP schemes~\cite{Starnberger2009, LeeHyunLim2010}, being them the basis for DAP\@. Still, no formal verification was done in any of these approaches. Therefore, there is no incontestable guarantee that they will not fail. Allied with a set of automated tools, formal verification can provide mathematical proofs that security protocols are correct or not. For that reason, we argue that formally verifying such critical systems is not only a valid motivation, but an obligation.



\section{Content Outline}

This document structure is divided in the following way:

\begin{enumerate}
    \item \textbf{Chapter~\ref{chap:sec-protocols}} presents the general theory in security protocols, discussing its properties and goals;

    \item \textbf{Chapter~\ref{chap:inductive}} begins with a brief introduction on formal verification and then describes our selected approach, the Inductive Method;

    \item \textbf{Chapter~\ref{chap:dap}} outlines internet banking systems and its challenges for presenting our targeted protocol, DAP, detailing its models, assumptions and operation;
    
    \item \textbf{Chapter~\ref{chap:formal-phones}} presents the formal model built for reasoning about smartphones and its operation. Also, it discusses another extensions done in the Inductive Method in order to provide a reliable model for the protocol formalization itself;

    \item \textbf{Chapter~\ref{chap:verification}} focus on the formalization and verification of the DAP itself. The security premises presented in the original specification of the DAP are used, following a more cautious approach. Many results are presented in this chapter; 
    
    \item \textbf{Chapter~\ref{chap:conclusion}} concludes this document, reviewing the results and discussing challenges and future work
\end{enumerate}
